---
title: "Fig5 CGI profile"
format:
  html:
    self-contained: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# Setup

```{r setup}
library(data.table)
library(dplyr)
library(purrr)
library(stringr)
library(ggplot2)
library(tibble)
library(readr)
library(ggh4x)

project_dir <- "/home/xuyanbo/project/MOF_TAPS"

knitr::opts_chunk$set(dpi = 300, fig.width = 6.5, fig.height = 3, dev = "cairo_pdf", dev.args = list(family = "Arial"))

font_family <- "Arial"

cgi_dir <- file.path(project_dir, "analysis", "fig_factory", "fig", "Fig5", "cgi")
methylation_dir <- file.path(cgi_dir, "methylation_bins")
coverage_dir <- file.path(cgi_dir, "coverage_bins")
coverage_dir <- file.path(cgi_dir, "coverage_bins")

out_table_dir <- cgi_dir
out_fig_dir <- cgi_dir

dir.create(out_table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(out_fig_dir, recursive = TRUE, showWarnings = FALSE)
```

# Sample metadata

```{r sample-meta}
input_order <- c("1 ng", "100 pg", "10 pg")
method_order <- c("MOF-TAPS", "TAPS", "TAPS+")

meta <- tibble(
  sample = c(
    "fym260106_S01", "fym260106_S02", "fym260106_S03",
    "fym260106_S04", "fym260106_S05", "fym260106_S06",
    "fym260124_S04", "fym260124_S05", "fym260124_S06"
  ),
  input_amount = factor(
    c(
      "10 pg", "100 pg", "1 ng",
      "10 pg", "100 pg", "1 ng",
      "10 pg", "100 pg", "1 ng"
    ),
    levels = input_order
  ),
  method = factor(
    c(
      "TAPS", "TAPS", "TAPS",
      "TAPS+", "TAPS+", "TAPS+",
      "MOF-TAPS", "MOF-TAPS", "MOF-TAPS"
    ),
    levels = method_order
  )
)

meta
```

# Theme

```{r theme}
method_colors <- c(
  "MOF-TAPS" = "#E59A2F",
  "TAPS+" = "#3B82C4",
  "TAPS" = "#63B34C"
)

theme_cgi <- function(show_legend = TRUE) {
  p <- theme_classic(base_size = 14, base_family = font_family) +
    theme(
      panel.border = element_blank(),
      axis.line = element_line(color = "black", linewidth = 0.4),
      axis.ticks = element_line(color = "black", linewidth = 0.4),
      axis.ticks.length = unit(0.16, "cm"),
      axis.text.x = element_text(size = 14, color = "black", margin = margin(t = 4), hjust = 1, vjust = 1),
      axis.text.y = element_text(size = 14, color = "black"),
      axis.title.x = element_text(size = 14, color = "black", margin = margin(t = 8)),
      axis.title.y = element_text(size = 14, color = "black", margin = margin(r = 8)),
      strip.background = element_blank(),
      strip.text = element_text(size = 14, color = "black", margin = margin(b = 8)),
      legend.position = "right",
      legend.title = element_blank(),
      legend.text = element_text(size = 14, color = "black", margin = margin(l = 2)),
      legend.key.height = unit(0.26, "cm"),
      legend.key.width = unit(0.65, "cm"),
      legend.key = element_blank(),
      legend.background = element_blank(),
      legend.box.background = element_blank(),
      plot.margin = margin(10, 10, 10, 10),
      panel.spacing.x = unit(1.2, "lines")
    )

  if (!show_legend) {
    p <- p + theme(legend.position = "none")
  }

  p
}
```

# Read CGI methylation bins

The methylation bin tables are expected to come from `bedtools map` on the CGI profile bins and final filtered TAPS-direction bedGraph. With a 7-column CGI-bin BED, the output should have 11 columns:

``` text
chrom  start  end  cgi_id  region  bin_index  profile_bin  mean_pct  sum_positive  sum_total  cpg_count
```

```{r read-methylation-bins}
read_methylation_bins <- function(sample) {
  f <- file.path(methylation_dir, paste0(sample, ".CGI_bins.methylation.tsv"))

  data.table::fread(f, header = FALSE, sep = "\t", na.strings = c(".", "NA")) %>%
    as_tibble() %>%
    transmute(
      sample = sample,
      chrom = V1,
      start = as.integer(V2),
      end = as.integer(V3),
      cgi_id = V4,
      region = V5,
      bin_index = as.integer(V6),
      profile_bin = as.integer(V7),
      mean_taps_readout_pct = as.numeric(V8),
      taps_positive = as.numeric(V9),
      total_informative = as.numeric(V10),
      cpg_count = as.numeric(V11),
      weighted_taps_readout_pct = if_else(
        !is.na(total_informative) & total_informative > 0,
        100 * taps_positive / total_informative,
        NA_real_
      )
    )
}

methylation_bin_df <- map_dfr(meta$sample, read_methylation_bins) %>%
  left_join(meta, by = "sample") %>%
  mutate(
    input_amount = factor(input_amount, levels = input_order),
    method = factor(method, levels = method_order)
  )

methylation_bin_df
```

# Read CGI coverage bins

The coverage bin tables are expected to come from per-base depth and `bedtools map`. With a 7-column CGI-bin BED, the output should have 8 columns:

``` text
chrom  start  end  cgi_id  region  bin_index  profile_bin  coverage_depth
```

```{r read-coverage-bins}
read_coverage_bins <- function(sample) {
  f <- file.path(coverage_dir, paste0(sample, ".CGI_bins.coverage.tsv"))

  x <- fread(
    f,
    sep = "\t",
    header = FALSE,
    col.names = c(
      "chrom", "start", "end", "cgi_id", "region", "bin_index", "profile_bin",
      "coverage_depth"
    )
  )

  x %>%
    as_tibble() %>%
    mutate(
      sample = sample,
      start = as.integer(start),
      end = as.integer(end),
      bin_index = as.integer(bin_index),
      profile_bin = as.integer(profile_bin),
      coverage_depth = as.numeric(coverage_depth)
    ) %>%
    select(sample, everything())
}

coverage_bin_df <- map_dfr(meta$sample, read_coverage_bins) %>%
  left_join(meta, by = "sample") %>%
  mutate(
    input_amount = factor(input_amount, levels = input_order),
    method = factor(method, levels = method_order)
  )

coverage_bin_df
```

# Summarize CGI profiles

For the methods-like profile, each CGI-bin value is first computed by `bedtools map`, then values from the same relative `profile_bin` are averaged across all CGI bins. For methylation, `mean_taps_readout_pct` is the direct mean of CpG-level TAPS readout within each CGI bin; `weighted_taps_readout_pct` is retained as a read-count-weighted alternative.

```{r summarize-profiles}
methylation_profile_df <- methylation_bin_df %>%
  group_by(input_amount, method, profile_bin) %>%
  summarise(
    n_bins = sum(!is.na(mean_taps_readout_pct)),
    mean_taps_readout_pct = mean(mean_taps_readout_pct, na.rm = TRUE),
    weighted_taps_readout_pct = mean(weighted_taps_readout_pct, na.rm = TRUE),
    mean_cpg_count = mean(cpg_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(input_amount, method, profile_bin)

coverage_profile_df <- coverage_bin_df %>%
  group_by(input_amount, method, profile_bin) %>%
  summarise(
    n_bins = sum(!is.na(coverage_depth)),
    mean_coverage_depth = mean(coverage_depth, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(input_amount, method, profile_bin)

write.csv(
  methylation_profile_df,
  file.path(out_table_dir, "HeLa_CGI_methylation_profile.csv"),
  row.names = FALSE
)

write.csv(
  coverage_profile_df,
  file.path(out_table_dir, "HeLa_CGI_coverage_profile.csv"),
  row.names = FALSE
)

methylation_profile_df
coverage_profile_df
```

```{r}
methylation_profile_df <- fread(file.path(out_table_dir, "HeLa_CGI_methylation_profile.csv"))

coverage_profile_df <- fread(file.path(out_table_dir, "HeLa_CGI_coverage_profile.csv"))
```

# Plot helper

```{r plot-helper}
profile_x_breaks <- c(1, 50.5, 70.5, 120)
profile_x_labels <- c("", "", "", "")

profile_label_df <- data.frame(
  profile_bin = c(1, 60.5, 120),
  y = -Inf,
  label = c("-4 kb", "CGI", "+4 kb"),
  hjust = c(0, 0.5, 1)
)

plot_cgi_profile <- function(plot_df,
                             y_col,
                             y_label,
                             show_legend = TRUE,
                             y_limits = NULL,
                             panel_y_limits = NULL,
                             panel_y_breaks = NULL,
                             y_breaks = waiver()) {
  plot_df <- plot_df %>%
    mutate(
      input_amount = factor(input_amount, levels = input_order),
      method = factor(method, levels = method_order)
    )

  get_panel_breaks <- function(panel_name) {
    if (is.null(panel_y_breaks)) {
      return(y_breaks)
    }

    if (!panel_name %in% names(panel_y_breaks)) {
      return(y_breaks)
    }

    panel_y_breaks[[panel_name]]
  }

  p <- ggplot(
    plot_df,
    aes(x = profile_bin, y = .data[[y_col]], color = method, group = method)
  ) +
    geom_line(linewidth = 0.75, na.rm = TRUE) +
    geom_vline(
      xintercept = c(50.5, 70.5),
      linetype = "dashed",
      linewidth = 0.35,
      color = "grey40"
    ) +
    geom_text(
      data = profile_label_df,
      aes(
        x = profile_bin,
        y = y,
        label = label,
        hjust = hjust
      ),
      inherit.aes = FALSE,
      vjust = 2.2,
      size = 4.8,
      color = "black"
    ) +
    facet_wrap(~ input_amount, nrow = 1, scales = "free_y") +
    scale_color_manual(
      values = method_colors,
      breaks = method_order,
      limits = method_order,
      drop = FALSE,
      name = NULL
    ) +
    guides(
      color = guide_legend(
        nrow = 1,
        byrow = TRUE
      )
    ) +
    scale_x_continuous(
      breaks = profile_x_breaks,
      labels = profile_x_labels,
      limits = c(1, 120),
      expand = expansion(mult = c(0.01, 0.01))
    ) +
    scale_y_continuous(
      limits = function(lims) {
        c(0, floor(lims[2]) + 1)
      },
      breaks = y_breaks,
      expand = expansion(mult = c(0, 0))
    )

  if (!is.null(panel_y_limits)) {
    p <- p +
      ggh4x::facetted_pos_scales(
        y = lapply(names(panel_y_limits), function(panel_name) {
          rlang::new_formula(
            lhs = rlang::expr(input_amount == !!panel_name),
            rhs = scale_y_continuous(
              limits = panel_y_limits[[panel_name]],
              breaks = get_panel_breaks(panel_name),
              expand = expansion(mult = c(0, 0))
            )
          )
        })
      )
  }

  p <- p +
    labs(x = NULL, y = y_label) +
    theme_cgi(show_legend = show_legend) +
    theme(
      plot.margin = margin(10, 10, 10, 10),
      legend.position = if (show_legend) "top" else "none",
      legend.justification = "center",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      legend.box.just = "center",
      legend.margin = margin(2, 2, 2, 2),
      legend.box.margin = margin(0, 0, 0, 0),
      panel.spacing.x = unit(0.4, "cm")
    )

  if (is.null(panel_y_limits) && !is.null(y_limits)) {
    p <- p + coord_cartesian(ylim = y_limits, clip = "off")
  } else {
    p <- p + coord_cartesian(clip = "off")
  }

  p
}
```

# Methylation level around CGI

```{r plot-methylation}
p_cgi_methylation <- plot_cgi_profile(
  methylation_profile_df,
  y_col = "mean_taps_readout_pct",
  y_label = "Average methylation level (%)",
  show_legend = TRUE,
  panel_y_limits = list(
    "1 ng" = c(0,80),
    "100 pg" = c (0,80),
    "10 pg" = c(0,80)
  )
)

p_cgi_methylation
```

```{r save-methylation}
ggsave(
  file.path(out_fig_dir, "HeLa_CGI_methylation_level_profile.pdf"),
  p_cgi_methylation,
  width = 8.7,
  height = 3.4,
  device = cairo_pdf
)
```

# Coverage depth around CGI

```{r plot-coverage}
p_cgi_coverage <- plot_cgi_profile(
  coverage_profile_df,
  y_col = "mean_coverage_depth",
  y_label = "Average coverage",
  show_legend = TRUE,
  panel_y_limits = list(
    "1 ng" = c(0, 15),
    "100 pg" = c (0, 8),
    "10 pg" = c(0, 2)
  ),
  panel_y_breaks = list(
    "1 ng" = seq(0, 15, 3),
    "100 pg" = seq(0, 8, 2),
    "10 pg" = seq(0, 2, 0.5)
  ),  
)

p_cgi_coverage
```

```{r save-coverage}
ggsave(
  file.path(out_fig_dir, "HeLa_CGI_coverage_depth_profile.pdf"),
  p_cgi_coverage,
  width = 8.7,
  height = 3.4,
  device = cairo_pdf
)
```

